A Formal Interface for Concurrent Search Structure Templates

### 2025-04-28 concurrency/template paper/paper.tex
outlined interfaces

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Abs}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Ref }}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\inFP}{\ensuremath{\mathsf{In }}}

@@

However, while \citet{templates} lay out the high-level ideas of the template approach and work through several examples, they do not formally define the approach. In particular, a modularization technique should have a well-defined \emph{interface} for each of its components, so that we know that components written and verified according to the technique can be successfully combined. In theory, given $m$ data structure components and $n$ concurrency templates, we should immediately be able to obtain $m \times n$ verified concurrent data structures. The original paper by Krishna et al. does not realize this (nor does it claim to): it presents three verified templates and five verified data structures, but each data structure is combined with only one template. In fact, close examination of the examples shows that different templates use different and incompatible specifications for the data structure component. In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Coq, that guarantees $m \times n$ verified data structures from $m + n$ components. More specifically:
\begin{itemize}
\item We describe an approach to \textbf{implementing} data structures and synchronization mechanisms so that they can be freely combined. This is a surprisingly challenging task: ideally, the data structure implementations should not explicitly leave space for concurrency metadata like locks, and the concurrency mechanism should not come between pieces of the data structure. We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits}, though it can also be implemented in C.
\item We give a formal \textbf{interface} for data structure components and concurrency templates, in the form of separation logic specifications for each of the relevant functions. These specifications follow the style of \citet{templates}, but we clarify which specifications should be considered the official interface and which are derived or implementation-specific, as well as ensuring that data structure components are never assumed to include concurrent features and concurrency templates never rely on data structure details. We verify the top-level concurrent data structure operations (\lstinline{insert} and \lstinline{lookup}) against these specifications, guaranteeing that they work correctly for any data structure component and template that satisfy the interface.
\item We exhibit two data structures (linked list and BST) and two concurrency templates (lock coupling and give-up) that satisfy our interface, and from them \textbf{freely obtain four verified concurrent data structures}, implemented in C and verified in VST. %The BST is particularly interesting because...
\end{itemize}

@@

The verification mechanism for search structure mechanisms is concurrent separation logic, and the interface between data structures and templates relies on two key constructs: \emph{flow interfaces} and \emph{logical atomicity}. Flow interfaces~\cite{krishna2017flow} are a generic mechanism for capturing the relationship of a single node to a larger data structure, and can be implemented in separation logic using ghost state. The interface provided by a data structure implementation includes a predicate of the form $\mathsf{node}\ n\ I\ C$, where $n$ is the node itself (i.e., a pointer to the node data structure), $I$ is $n$'s flow interface, and $C$ is $n$'s contribution to the state of the overall data structure, e.g., the set of keys contained in $n$. Data structure operations such as \lstinline{insert}, \lstinline{lookup}, and \lstinline{findNext} are specified in terms of the $\mathsf{node}$ predicate. %example?

@@

\subsection{Iris and VST}
The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions. Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications. This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs? or show them here?
Algorithms are verified in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little
}

@@

\section{Specifying the Template Interface}
Our verification process for concurrent search structure templates follows the same modular architecture as the code itself:
\begin{itemize}
\item The interface for data structures is an abstract $\mathsf{node}$ predicate, plus sequential Hoare triples for the data structure functions (\lstinline{findNext}, \lstinline{insertOp}, \lstinline{lookupOp}) that characterize their behavior at the level of $\mathsf{node}$s.
\item The interface for templates takes an arbitrary data structure as a parameter, and gives logically atomic triples for the template functions (\lstinline{traverse}, \lstinline{insertHelper}) in terms of their effects on an abstract $\mathsf{CSS}$ predicate.
\item The top-level \lstinline{insert} and \lstinline{lookup} functions are specified and verified once and for all, taking both a data structure and a template as a parameter, and providing proofs that the top-level \lstinline{insert} and \lstinline{lookup} functions satisfy atomic specifications on $\mathsf{CSS}$:
\[{\color{specblue}
		\fforall m.\left\langle 
		\mathsf{CSS}(m)
		\right\rangle
	}\ \texttt{insert(r, k, v)}\ 
	{\color{specblue}
		\left\langle 
		\mathsf{CSS}(m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}\]
\[{\color{specblue}
		\fforall m.\left\langle 
		\mathsf{CSS}(m)
		\right\rangle
	}\ \texttt{lookup(r, k)}\ 
	{\color{specblue}
		\left\langle v.\ 
		\mathsf{CSS}(m) \land m(k) = v
		\right\rangle
	}\]
\end{itemize}
In this section, we present the precise specifications for the data structure and template functions. The structure of the specifications guarantees compositionality: if we have $m$ data structures that satisfy the data structure interface, and $n$ templates that satisfy the template interface, we can freely combine them to get $m \times n$ verified concurrent search structures. As is typical when defining interfaces, the key challenge is to define specifications for the data structure functions that give enough information to verify any template, but are general enough to be satisfied by any data structure. ...

@@

\section{Conclusion and Future Work}
The framework we have presented fulfills the true promise of the concurrency template approach: each data structure instance is implemented and verified without any awareness of concurrency or metadata, each template instance is implemented and proved without any knowledge of the implementation of the data structure, and we obtain verified concurrent data structures immediately by instantiating our top-level theorems with any combination of data structure and template instances. We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Coq specification and proof. Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics). %We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.

\ignore{
\begin{acks}
We thank Nisarg Patel, ...
This work was partially funded by the National Science Foundation under the award ???.
\end{acks}
}

>>>>

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{Abs}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{Ref }}}
\newcommand{\lockinv}{\ensuremath{\mathsf{lock\_inv}}}
\newcommand{\inFP}{\ensuremath{\mathsf{In }}}
\newcommand{\node}[3]{\ensuremath{\mathsf{node}(#1, #2, #3)}}
\newcommand{\mdentry}[3]{\ensuremath{\mathsf{md\_node}(#1, #2, #3)}} %will need more arguments -- what are the arguments to md_entry_rep?

@@

However, while \citet{templates} lay out the high-level ideas of the template approach and work through several examples, they do not formally define the approach. In particular, a modularization technique should have a well-defined \emph{interface} for each of its components, so that we know that components written and verified according to the technique can be successfully combined. In theory, given $m$ data structure components and $n$ concurrency templates, we should immediately be able to obtain $m \times n$ verified concurrent data structures. The original paper by Krishna et al. does not realize this (nor does it claim to): it presents three verified templates and five verified data structures, but each data structure is combined with only one template. In fact, close examination of the examples shows that different templates use different and incompatible specifications for the data structure component. In this paper, we rectify this by presenting a \textbf{formal interface for concurrent search structure templates}, implemented as a library in C and a collection of typeclasses in Rocq, that guarantees $m \times n$ verified data structures from $m + n$ components. More specifically:
\begin{itemize}
\item We describe an approach to \textbf{implementing} data structures and synchronization mechanisms so that they can be freely combined. This is a surprisingly challenging task: ideally, the data structure implementations should not explicitly leave space for concurrency metadata like locks, and the concurrency mechanism should not come between pieces of the data structure. We discuss several undesirable approaches and their relationships to object-orientation patterns, and show that the most desirable approach corresponds to \emph{traits}, though it can also be implemented in C.
\item We give a formal \textbf{interface} for data structure components and concurrency templates, in the form of separation logic specifications for each of the relevant functions. These specifications follow the style of \citet{templates}, but we clarify which specifications should be considered the official interface and which are derived or implementation-specific, as well as ensuring that data structure components are never assumed to include concurrent features and concurrency templates never rely on data structure details. We verify the top-level concurrent data structure operations (\lstinline{insert} and \lstinline{lookup}) against these specifications, guaranteeing that they work correctly for any data structure component and template that satisfy the interface.
\item We exhibit two data structures (linked list and BST) and two concurrency templates (lock coupling and give-up) that satisfy our interface, and from them \textbf{freely obtain four verified concurrent data structures}, implemented in C and verified in VST. %The BST is particularly interesting because...
\end{itemize}

@@

The verification mechanism for search structure mechanisms is concurrent separation logic, and the interface between data structures and templates relies on two key constructs: \emph{flow interfaces} and \emph{logical atomicity}. Flow interfaces~\cite{krishna2017flow} are a generic mechanism for capturing the relationship of a single node to a larger data structure, and can be implemented in separation logic using ghost state. The interface provided by a data structure implementation includes a predicate of the form $\node{n}{I}{C}$, where $n$ is the node itself (i.e., a pointer to the node data structure), $I$ is $n$'s flow interface, and $C$ is $n$'s contribution to the state of the overall data structure, e.g., the set of keys contained in $n$. Data structure operations such as \lstinline{insert}, \lstinline{lookup}, and \lstinline{findNext} are specified in terms of the $\mathsf{node}$ predicate. %example? explain flow interfaces further?

@@

\subsection{Iris and VST}
The search structure template approach uses concurrent separation logic to specify and prove pre- and postconditions for the template and data structure functions. Krishna et al.~implemented their framework in Iris~\cite{iris}, a language-independent CSL framework built in the Rocq proof assistant~\cite{rocq}, with flexible support for ghost state, invariants, and atomic specifications. This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs? or show them here?
Algorithms are verified in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Rocq and is connected to the CompCert verified C compiler~\cite{compcert}, allowing it to guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{iris-vst-arxiv} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. In this paper, we use these extensions to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little
}

@@

\section{Specifying Data Structures and Templates as Interfaces}
Our verification process for concurrent search structure templates follows the same modular architecture as the code itself:
\begin{itemize}
\item The interface for data structures is an abstract $\mathsf{node}$ predicate, plus sequential Hoare triples for the data structure functions (\lstinline{findNext}, \lstinline{insertOp}, \lstinline{lookupOp}) that characterize their behavior at the level of $\mathsf{node}$s.
\item The interface for templates takes an arbitrary data structure as a parameter, and gives logically atomic triples for the template functions (\lstinline{traverse}, \lstinline{insertHelper}) in terms of their effects on an abstract $\mathsf{CSS}$ predicate.
\item The top-level \lstinline{insert} and \lstinline{lookup} functions are specified and verified once and for all, taking both a data structure and a template as a parameter, and providing proofs that the top-level \lstinline{insert} and \lstinline{lookup} functions satisfy atomic specifications on $\mathsf{CSS}$:
\[{\color{specblue}
		\fforall m.\left\langle 
		\mathsf{CSS}(m)
		\right\rangle
	}\ \texttt{insert(r, k, v)}\ 
	{\color{specblue}
		\left\langle 
		\mathsf{CSS}(m[\texttt{k} \mapsto \texttt{v}])
		\right\rangle
	}\]
\[{\color{specblue}
		\fforall m.\left\langle 
		\mathsf{CSS}(m)
		\right\rangle
	}\ \texttt{lookup(r, k)}\ 
	{\color{specblue}
		\left\langle v.\ 
		\mathsf{CSS}(m) \land m(k) = v
		\right\rangle
	}\]
\end{itemize}
In this section, we present the precise specifications for the data structure and template functions. The structure of the specifications guarantees compositionality: if we have $m$ data structures that satisfy the data structure interface, and $n$ templates that satisfy the template interface, we can freely combine them to get $m \times n$ verified concurrent search structures. As is typical when defining interfaces, the key challenge is to define specifications for the data structure functions that give enough information to verify any template, but are general enough to be satisfied by any data structure. Each of our interfaces is implemented as a typeclass in Rocq, which makes it easy to do proofs over a generic instance of the interface.

\subsection{Data Structure Interface}

The data structure interface is built around an abstract predicate $\node{n}{I}{C}$ representing a node in the data structure, where $n$ is the concrete pointer to the node, $I$ is the flow interface for the node (an abstraction of its position in the data structure), and $C$ is the contents of the note as a map from keys to values. In practice, $\mathsf{node}$ will be implemented as a combination of concrete points-to predicates for the physical node in memory, and ghost state representing the node's contribution to the abstract state of the data structure. %discuss how this relates to the acsys proofs?
We then specify the key data structure operations as follows: %fill these in
\begin{mathpar}
\texttt{findNext}

\texttt{insertOp}

\texttt{lookupOp}
\end{mathpar}

\subsection{Template Interface}
The template interface includes two abstract predicates: \treerep(p, C), which describes the abstract state of the entire concurrent data structure at pointer $p$ as a key-value map $C$, and $\mdentry{n}{I}{C}$, %more arguments
which represents the combination of a node and the metadata attached to it in the template. Prior work does not clearly distinguish between $\mathsf{node}$ and $\mathsf{md\_node}$---for instance, \citet{templates} assume that the $\mathsf{node}$ predicate includes the lock and, in the give-up template, the range for the node---but this distinction is necessary for truly modular specification: the data structure interface should not include concurrency metadata, since it is both concurrency-specific and different for each template instance. Thus, while data structure functions act on $\mathsf{node}$s, the template functions are defined entirely at the level of $\mathsf{md\_node}$s.

The template operations are specified as follows: %fill these in
\begin{mathpar}
\texttt{traverse}

\texttt{insertHelper}

\texttt{lookupHelper}
\end{mathpar}
Each top-level operation will be implemented with a combination of \texttt{traverse}, which traverses the data structure to find the $\mathsf{md\_node}$ where a key belongs, and one of the helper functions, which completes the chosen operation given the target $\mathsf{md\_node}$. The helper functions are responsible for both performing the actual data structure operation (using \texttt{insertOp} or \texttt{lookupOp} from the data structure interface) and updating the metadata (creating new locks, modifying ranges, etc.) to create a consistent concurrent data structure for the new state.

\subsection{Top-Level Operations}
Given instances of the data structure and template interfaces, we can define and verify the top-level \texttt{insert} and \texttt{lookup} operations for concurrent search structures. (implementations and proofs here)

In Rocq, these proofs are parameterized by instances of the $\mathsf{DataStructure}$ and $\mathsf{Template}$ classes. It is precisely this that gives us true modularity: we know that for \emph{any} data structure implementation meeting the data structure interface, and \emph{any} template meeting the template interface, we can instantiate these parameterized proofs and obtain verified \texttt{insert} and \texttt{lookup} functions. In other words, this guarantees that given $m$ data structure implementations and $n$ template implementations, we can immediately and with no additional effort obtain $m \times n$ verified concurrent data structures.

@@

\section{Conclusion and Future Work}
The framework we have presented fulfills the true promise of the concurrency template approach: each data structure instance is implemented and verified without any awareness of concurrency or metadata, each template instance is implemented and proved without any knowledge of the implementation of the data structure, and we obtain verified concurrent data structures immediately by instantiating our top-level theorems with any combination of data structure and template instances. We accomplished this by careful attention to which components belong to the data structure and which to the template, and by working to disentangle the two in both the C implementation and the Rocq specification and proof. Our interfaces are general enough that it should be easy to add more data structures and templates, including concurrency patterns that do not depend on locks (e.g., optimistic concurrency control with fine-grained atomics). %We have seen that the concurrency template must be implemented in a very specific way, with metadata stored separately from the data structure itself, in order to sufficiently separate concerns; in future work, we are interested to see whether languages with more flexible composition mechanisms, like traits in Rust and Scala, lend themselves to more natural implementations of this composition.

\ignore{
\begin{acks}
We thank Nisarg Patel for detailed discussions about the original search structure template proofs.
This work was partially funded by the National Science Foundation under the award ???.
\end{acks}
}
